Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change

### 2020-12-03 Scripts/Manuscript.Rmd
Changes for dealing with word

#### Project's Library
packages <- c(
  "tidyverse" ,
  "png",
  "grid",
  "knitr",
 # "kableExtra"
  # "here", # for dbem_import `here()`
  # "data.table", #dbem_import `fread()`
  # "readxl", # for reading excell files
  # "janitor", # for clearing names
  # "geosphere", # estimate distances between points `distm()`
  # "ggrepel",
  # "zoo", # for average mean
  # "parallel", # for mclapply,
  # "sf", # for mapping
  # "st", # for mapping
  # "rgdal", #Spatial analysis
  # "tools", #Spatial analysis 
  # "zeallot", # for Juanito's map
  # "gmt", # for estimating distances between points
  # "viridis"
)

@@

The average time of emergence of the transboundary index across the emerging stocks in all EEZs analyzed is projected to be 2029 \pm 26 years (Figure \@ref(fig:toeFigure1)). In total, the projected transboundary index of 5,745 stocks (63% of the studied stocks) will emerge from their historical variability between 2006 and 2100. Fish stocks with transboundary index that first emerged from their historical variability in each EEZ was projected to be as early as 2006 and the last one in 2100. About 55% (n = 3,154) of those emerging stocks are projected to have their transboundary index emerged between 2006 and 2020 (e.g. Peruvean anchoveta - *Engraulis ringens* - shared by Chile and Peru emerged in 2010). The number of EEZs with emerging stocks is projected to increase steeply from 87% in 2006 to 96% in 2020. After the mid-2020s, the rate of increase in the number EEZs slow down to an average of 0.7% per year. By 2081, all of the EEZs analyzed here will have at least one emerging stock (Figure \@ref(fig:toeFigure1)). Moreover, 96% of the world’s EEZs saw at least one stock having their distribution shifts between 2006 and 2020 and around one fifth (19%) of the EEZs have an average time of emergence of the transboundary index across stocks prior to 2020. Shifts in stocks’ distributions will continue to exceed historical limits steadily towards the end of the $21^{st}$ century (Figure Figure \@ref(fig:toeFigure1)). I based our transboundary index on a transboundary stock’s centroid defined as the average grid cells that present the higher stock’s abundance (see Methods - Calculating an index of transboundary range shift). Results from the sensitivity analysis suggest that our method of estimating the time of emergence of the transboundary index is robust to the abundance level (e.g. the top 90%) of the stock (Figure \@ref(fig:toeFigureS3).

@@

I set an arbitrary threshold of 1 s.d. (67% confidence) to determine the time of emergence of the transboundary index and the change in stock share ratio below the threat point (see Methods). I tested the sensitivity of our methods by considering a larger threshold for the time of emergence (e.g., 2 s.d for a 95% confidence). The global average time of emergence of the transboundary index considering a 2 s.d threshold was 2036 \pm 28, a difference of about 7 years relative to the 1 s.d. threshold (Figure \@ref(fig:toeFigureS2)). In total, 4,119 transboundary stocks would emerge between 2006 and 2100 considering a more conservative metric, that is, 40% (n = 1626) fewer emerging stocks than using a 1.s.d threshold.

@@

I have identified the stocks that are projected to have changes in stock share ratio beyond an EEZ’s threat point and estimated the changes in the stock share ratio by the early (2020-2040) and the mid $21^{st}$ century (2040-2060), relative to the recent past (1951-2005) (see methods - equation \@ref(eq:five)). By the early $21^{st}$ century, the global average proportion of transboundary stocks that are projected to change their stock share ratio beyond the EEZ’s threat point is 18% \pm 15% (Figure \@ref(fig:toeFigure5)A). The changes in stock share ratio are expected to increase slightly towards the mid $21^{st}$ century (Figures \@ref(fig:toeFigureS4) and \@ref(fig:toeFigureS5)). However, changes in stock share ratio for specific stocks and in some EEZs are projected to be exceptionally high. For example, Guatemala shares Panulirus gracilis stock with Mexico, and Guatemala’s share is projected to increase by >50%, that is from 12% to 26% by the early $21^{st}$ century. Towards the mid $21^{st}$ century, changes in stock share ratio per EEZ will not be substantial (Figure \@ref(fig:toeFigureS5)) relative to the early $21^{st}$ century but are expected to increase in intensity (Figure \@ref(fig:toeFigureS4)). Our results suggest that 87% (n = 238) of the world EEZs will experience changes in stock share ratio beyond a country’s threat point by the early $21^{st}$ century (Figure \@ref(fig:toeFigure5)A). These EEZs will see one third (33% \pm 21%) of their transboundary stocks changing the stock share ratio beyond the threat point (Figure \@ref(fig:toeFigure5)A). Particularly, Brazil, French Guiana, Guam, Kergelen and Pitcairin will see changes in all of the transboundary stocks analyzed while Venezuela (6%), Bonaire (9%), China (10%) and Vietnam (10%) will be the only countries with less than 10% of stocks changing beyond the threat point.

@@

The presentce of EEZs lacking stocks with emerging transboundary index is a combination of EEZ shape, climate projections, and a reduced number of transbounudary stocks. For example, there are 59 commercial fish stocks fished the Arctic with the vast majority focused on the Barents, Bering and Norwegian seas with limited knowledge about the species biology [@Christiansen:2014gg]. However, the Arctic has been warming faster than many other regions resulting in a reduction in sea-ice [@Douglas:2010Mb] that could potentially be occupied by species previously not present [@REF]. As a result, the biomass of Arctic marine animals [@BryndumBuchholz:2019kh] and catch and revenue of associated fisheries [@Tai:2019cq; @Lam:2014fq] are expected to increase in the incoming years, although substantial uncertainty exists on the magnitude of change [@BryndumBuchholz:2019kh]. Moreover, projections suggest a larger increment on transboundary stocks in the incoming years in the Arctic [@Pinsky:2018cb] as climate change pushes sub-polar species further North [@Morley:2018fn; @Frainer:2017ji]. In the Case of Brazil, in addition to having few transbounudary stocks (only 4 shared with Uruguay in this study), the large diffrence in the EEZs lenghts could also be influencing the results.

@@

Changes in stock share ratio beyond the threat point of countries sharing a common resource could instigate international conflict and hinder bi-lateral agreements (@Miller:2013iv; @Sumaila:2020tp). Lessons learned from historic distributional shifts of shared stocks suggest that treaties that are not prepare to respond to such shifts will be less resilient to changes in transboundary fish stocks share ratio due to climate change [@Miller:2013iv; @Sumaila:2020tp]. For example, disputes over Pacific salmon between Canada and the USA arose in the 1990s when a climate-related shift in stock abundance favoured Alaska’s salmon fisheries over Canada’s [@Miller:2013iv; @Munro:2002uf; @Song:2017va]. The conflict lasted over ten years until Canada and the USA agreed on a mutual conservation fund where the US would contribute the larger proportion to support scientific research, habitat restoration and enhancement of wild stock production in transboundary rivers [@Munro:2002uf]. Identified strategies to cope with changes in the share proportion of transboundary stocks include the incorporation of side payments [@Tunca:2019if], increased international cooperation [@Miller:2013iv] and dynamic ocean management [@Chapter5; @Pinsky2018]. Side payments are mechanisms that provide a solution to the inequality produced by shifts in the stock’s distribution [@Tunca:2019if; @Miller:2004fpa]. These can be monetary (e.g., the stock-“winning” State compensates the stock-“loosing” state for the proportional shift) or other methods like the Pacific salmon case (e.g., a conservation fund payed mainly by one Party) [@Miller:2004fpa; @Miller:2013iv].  Strengthening current international cooperation will be required as stock distribution will move to neighboring jurisdictions, requiring combine efforts to generate information and set proper management rules [@Miller:2013iv]. Finally, flexible management rules might be better fitted to capture the stock’s distributional shift. For example, quota allocation based on the stock’s current distribution like that for Pacific halibut (*Hippoglossus stenolepis*) managed by Canada and the US [@Sumaila:2020tp; @PalaciosAbrantes:2020vt] might be better able to capture the stock’s shifts than a quota based on a fix-historical proportion like the case of the EU [@Baudron:2020dj].

@@

```{r toeFigureS1, eval = F,  echo = F, fig.width=9, fig.height=4, fig.cap =  "World Exclusive Economic Zones used in this study as devided by the Sea Around Us and their centroids (points)."}

@@

```{r toeFigureS2, eval = F,  echo = F, fig.width=9, fig.height=4, fig.cap = "Graphical representation of the transboundary index (TI) to determine the time of emergence of transboundary stocks. The index is based on the distance between the distributional centroid of the transboundary stock and the geographic centroid of the neighbouring Exclusive Economic Zones sharing the stock (Top panel). Time of emergence is defined as the first year when the TI overpasses historical values (Bottom panel)."}

@@

```{r toeFigureS3, eval = F,  echo = F, fig.width=4, fig.height=4, fig.cap =  "Average Time  of Emergence and standar dviation for a 64% confidence treshold  (purple) and a 98% confidence teshold (yellow)"}

@@

```{r toeFigureS4, eval = F,  echo = F, fig.width=4, fig.height=4, fig.cap =  "Distribution of Stock Share Ratio percentage change by the early (2021-2040) and middle (2040-2060) $21^{st}$ centurury relative to today (1951-2005)"}

>>>>

#### Project's Library
packages <- c(
  "tidyverse" ,
  "png",
  "grid",
  "knitr"#,
 # "kableExtra"
  # "here", # for dbem_import `here()`
  # "data.table", #dbem_import `fread()`
  # "readxl", # for reading excell files
  # "janitor", # for clearing names
  # "geosphere", # estimate distances between points `distm()`
  # "ggrepel",
  # "zoo", # for average mean
  # "parallel", # for mclapply,
  # "sf", # for mapping
  # "st", # for mapping
  # "rgdal", #Spatial analysis
  # "tools", #Spatial analysis 
  # "zeallot", # for Juanito's map
  # "gmt", # for estimating distances between points
  # "viridis"
)

@@

The average time of emergence of the transboundary index across the emerging stocks in all EEZs analyzed is projected to be 2029 $\pm$ 26 years (Figure \@ref(fig:toeFigure1)). In total, the projected transboundary index of 5,745 stocks (63% of the studied stocks) will emerge from their historical variability between 2006 and 2100. Fish stocks with transboundary index that first emerged from their historical variability in each EEZ was projected to be as early as 2006 and the last one in 2100. About 55% (n = 3,154) of those emerging stocks are projected to have their transboundary index emerged between 2006 and 2020 (e.g. Peruvean anchoveta - *Engraulis ringens* - shared by Chile and Peru emerged in 2010). The number of EEZs with emerging stocks is projected to increase steeply from 87% in 2006 to 96% in 2020. After the mid-2020s, the rate of increase in the number EEZs slow down to an average of 0.7% per year. By 2081, all of the EEZs analyzed here will have at least one emerging stock (Figure \@ref(fig:toeFigure1)). Moreover, 96% of the world’s EEZs saw at least one stock having their distribution shifts between 2006 and 2020 and around one fifth (19%) of the EEZs have an average time of emergence of the transboundary index across stocks prior to 2020. Shifts in stocks’ distributions will continue to exceed historical limits steadily towards the end of the $21^{st}$ century (Figure Figure \@ref(fig:toeFigure1)). I based our transboundary index on a transboundary stock’s centroid defined as the average grid cells that present the higher stock’s abundance (see Methods - Calculating an index of transboundary range shift). Results from the sensitivity analysis suggest that our method of estimating the time of emergence of the transboundary index is robust to the abundance level (e.g. the top 90%) of the stock (Figure \@ref(fig:toeFigureS3).

@@

I set an arbitrary threshold of 1 s.d. (67% confidence) to determine the time of emergence of the transboundary index and the change in stock share ratio below the threat point (see Methods). I tested the sensitivity of our methods by considering a larger threshold for the time of emergence (e.g., 2 s.d for a 95% confidence). The global average time of emergence of the transboundary index considering a 2 s.d threshold was 2036 $\pm$ 28, a difference of about 7 years relative to the 1 s.d. threshold (Figure \@ref(fig:toeFigureS2)). In total, 4,119 transboundary stocks would emerge between 2006 and 2100 considering a more conservative metric, that is, 40% (n = 1626) fewer emerging stocks than using a 1.s.d threshold.

@@

I have identified the stocks that are projected to have changes in stock share ratio beyond an EEZ’s threat point and estimated the changes in the stock share ratio by the early (2020-2040) and the mid $21^{st}$ century (2040-2060), relative to the recent past (1951-2005) (see methods - equation \@ref(eq:five)). By the early $21^{st}$ century, the global average proportion of transboundary stocks that are projected to change their stock share ratio beyond the EEZ’s threat point is 18% $\pm$ 15% (Figure \@ref(fig:toeFigure5)A). The changes in stock share ratio are expected to increase slightly towards the mid $21^{st}$ century (Figures \@ref(fig:toeFigureS4) and \@ref(fig:toeFigureS5)). However, changes in stock share ratio for specific stocks and in some EEZs are projected to be exceptionally high. For example, Guatemala shares Panulirus gracilis stock with Mexico, and Guatemala’s share is projected to increase by >50%, that is from 12% to 26% by the early $21^{st}$ century. Towards the mid $21^{st}$ century, changes in stock share ratio per EEZ will not be substantial (Figure \@ref(fig:toeFigureS5)) relative to the early $21^{st}$ century but are expected to increase in intensity (Figure \@ref(fig:toeFigureS4)). Our results suggest that 87% (n = 238) of the world EEZs will experience changes in stock share ratio beyond a country’s threat point by the early $21^{st}$ century (Figure \@ref(fig:toeFigure5)A). These EEZs will see one third (33% $\pm$ 21%) of their transboundary stocks changing the stock share ratio beyond the threat point (Figure \@ref(fig:toeFigure5)A). Particularly, Brazil, French Guiana, Guam, Kergelen and Pitcairin will see changes in all of the transboundary stocks analyzed while Venezuela (6%), Bonaire (9%), China (10%) and Vietnam (10%) will be the only countries with less than 10% of stocks changing beyond the threat point.

@@

The presentce of EEZs lacking stocks with emerging transboundary index is a combination of EEZ shape, climate projections, and a reduced number of transbounudary stocks. For example, there are 59 commercial fish stocks fished the Arctic with the vast majority focused on the Barents, Bering and Norwegian seas with limited knowledge about the species biology [@Christiansen:2014gg]. However, the Arctic has been warming faster than many other regions resulting in a reduction in sea-ice [@Douglas:2010Mb] that could potentially be occupied by species previously not present [@IPCC:2019tn]. As a result, the biomass of Arctic marine animals [@BryndumBuchholz:2019kh] and catch and revenue of associated fisheries [@Tai:2019cq; @Lam:2014fq] are expected to increase in the incoming years, although substantial uncertainty exists on the magnitude of change [@BryndumBuchholz:2019kh]. Moreover, projections suggest a larger increment on transboundary stocks in the incoming years in the Arctic [@Pinsky:2018cb] as climate change pushes sub-polar species further North [@Morley:2018fn; @Frainer:2017ji]. In the Case of Brazil, in addition to having few transbounudary stocks (only 4 shared with Uruguay in this study), the large diffrence in the EEZs lenghts could also be influencing the results.

@@

Changes in stock share ratio beyond the threat point of countries sharing a common resource could instigate international conflict and hinder bi-lateral agreements (@Miller:2013iv; @Sumaila:2020tp). Lessons learned from historic distributional shifts of shared stocks suggest that treaties that are not prepare to respond to such shifts will be less resilient to changes in transboundary fish stocks share ratio due to climate change [@Miller:2013iv; @Sumaila:2020tp]. For example, disputes over Pacific salmon between Canada and the USA arose in the 1990s when a climate-related shift in stock abundance favoured Alaska’s salmon fisheries over Canada’s [@Miller:2013iv; @Munro:2002uf; @Song:2017va]. The conflict lasted over ten years until Canada and the USA agreed on a mutual conservation fund where the US would contribute the larger proportion to support scientific research, habitat restoration and enhancement of wild stock production in transboundary rivers [@Munro:2002uf]. Identified strategies to cope with changes in the share proportion of transboundary stocks include the incorporation of side payments [@Tunca:2019if], increased international cooperation [@Miller:2013iv] and dynamic ocean management [@Pinsky:2018cb, Chapter 5]. Side payments are mechanisms that provide a solution to the inequality produced by shifts in the stock’s distribution [@Tunca:2019if; @Miller:2004fpa]. These can be monetary (e.g., the stock-“winning” State compensates the stock-“loosing” state for the proportional shift) or other methods like the Pacific salmon case (e.g., a conservation fund payed mainly by one Party) [@Miller:2004fpa; @Miller:2013iv].  Strengthening current international cooperation will be required as stock distribution will move to neighboring jurisdictions, requiring combine efforts to generate information and set proper management rules [@Miller:2013iv]. Finally, flexible management rules might be better fitted to capture the stock’s distributional shift. For example, quota allocation based on the stock’s current distribution like that for Pacific halibut (*Hippoglossus stenolepis*) managed by Canada and the US [@Sumaila:2020tp; @PalaciosAbrantes:2020vt] might be better able to capture the stock’s shifts than a quota based on a fix-historical proportion like the case of the EU [@Baudron:2020dj].

@@

```{r FigureS1, eval = F,  echo = F, fig.width=9, fig.height=4, fig.cap =  "World Exclusive Economic Zones used in this study as devided by the Sea Around Us and their centroids (points)."}

@@

```{r FigureS2, eval = F,  echo = F, fig.width=9, fig.height=4, fig.cap = "Graphical representation of the transboundary index (TI) to determine the time of emergence of transboundary stocks. The index is based on the distance between the distributional centroid of the transboundary stock and the geographic centroid of the neighbouring Exclusive Economic Zones sharing the stock (Top panel). Time of emergence is defined as the first year when the TI overpasses historical values (Bottom panel)."}

@@

```{r FigureS3, eval = F,  echo = F, fig.width=4, fig.height=4, fig.cap =  "Average Time  of Emergence and standar dviation for a 64% confidence treshold  (purple) and a 98% confidence teshold (yellow)"}

@@

```{r FigureS4, eval = F,  echo = F, fig.width=4, fig.height=4, fig.cap =  "Distribution of Stock Share Ratio percentage change by the early (2021-2040) and middle (2040-2060) $21^{st}$ centurury relative to today (1951-2005)"}

### 2020-12-03 Scripts/Supplements.Rmd
Changes for dealing with word

```{r setup, eval = T, echo=F, warning=F,message=F, results='hide'}
library(MyFunctions)
#### Library ####
packages <- c(
  "dplyr", # Data manipulation
  "knitr",
  "kableExtra",
  "png",
  "grid",
  "janitor"
)
my_lib(packages)
```

@@

read.csv("~/GitHub/TransEmergence/Tables/kw_regional.csv") %>% 
  select(-statistic,
         "Habitat preference comparrison" = comparrison,
         "Significan level" = signif.level,
         "Observation difference" = dif.com.obs.dif,
         "Critical difference" = dif.com.critical.dif,
          "K-W Difference" = dif.com.difference) %>% 
  kable("latex",
        booktabs = T,
        caption = "Statistical results for multiple comparison test after Kruskal-Wallis on time of emergence across United Nations sub regions"
  ) %>% 
  kable_styling(full_width = F,
                position = "center")

@@

read.csv("~/GitHub/TransEmergence/Tables/kw_habitat.csv") %>% 
  select(-statistic,
         "Sub region comparrison" = comparrison,
         "Significan level" = signif.level,
         "Observation difference" = dif.com.obs.dif,
         "Critical difference" = dif.com.critical.dif,
          "K-W Difference" = dif.com.difference) %>% 
  kable("latex",
        booktabs = T,
        caption = "Statistical results for multiple comparison test after Kruskal-Wallis on time of emergence across United Nations sub regions"
  ) %>% 
  kable_styling(full_width = F,
                position = "center")

>>>>

```{r setup, eval = T, echo=F, warning=F,message=F, results='hide'}
library(MyFunctions)
#### Library ####
packages <- c(
  "dplyr", # Data manipulation
  "knitr",
  # "kableExtra",
  "png",
  "grid",
  "janitor"
)
my_lib(packages)
```

@@

read.csv("~/GitHub/TransEmergence/Tables/kw_regional.csv") %>% 
  select(-statistic,
         "Habitat preference comparrison" = comparrison,
         "Significan level" = signif.level,
         "Observation difference" = dif.com.obs.dif,
         "Critical difference" = dif.com.critical.dif,
          "K-W Difference" = dif.com.difference) %>% 
  kable(caption = "Statistical results for multiple comparison test after Kruskal-Wallis on time of emergence across United Nations sub regions")
  # kable("latex",
        # booktabs = T,
        # caption = "Statistical results for multiple comparison test after Kruskal-Wallis on time of emergence across United Nations sub regions"
  # ) %>% 
  # kable_styling(full_width = F,
                # position = "center")

@@

read.csv("~/GitHub/TransEmergence/Tables/kw_habitat.csv") %>% 
  select(-statistic,
         "Sub region comparrison" = comparrison,
         "Significan level" = signif.level,
         "Observation difference" = dif.com.obs.dif,
         "Critical difference" = dif.com.critical.dif,
          "K-W Difference" = dif.com.difference) %>% 
  kable(
        booktabs = T,
        caption = "Statistical results for multiple comparison test after Kruskal-Wallis on time of emergence across United Nations sub regions"
  )  
  # kable("latex",
  #       booktabs = T,
  #       caption = "Statistical results for multiple comparison test after Kruskal-Wallis on time of emergence across United Nations sub regions"
  # ) %>% 
  # kable_styling(full_width = F,
  #               position = "center")
